---
title: "Predicting Mortgage Yield using Regression Analysis"
author: "Group 42"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    number_sections: true
    toc: false
    latex_engine: pdflatex
    keep_tex: true
geometry: margin=2.5cm
fontsize: 11pt
header-includes:
  - \usepackage{graphicx}
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \usepackage{fancyhdr}
  - \usepackage{ragged2e}
  - \usepackage{multicol}
  - \justifying
  - \pagestyle{fancy}
  - \fancyhead[L]{Delandre, Garcia, Biocchi, Leteurtre}
  - \fancyfoot[C]{\thepage}
  - \usepackage{titlesec}
  - \titlespacing*{\title}{0pt}{-2cm}{0pt}  # Move title higher
  - \titlespacing*{\author}{0pt}{-2cm}{0pt}  # Move author higher
  - \titlespacing*{\date}{0pt}{-2cm}{0pt}    # Move date higher
  - \setlength{\topskip}{0pt}  # Reduce space at top of page
  - \setlength{\headheight}{0pt}  # Remove extra header space
  - \setlength{\headsep}{0pt}  # Remove extra space between header and content

editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

The study of A. H. Schaaf, 1966, "Regional Differences in Mortgage
Financing Costs", investigates the existence and causes of regional
differences in Mortgage financing costs in the United States. While
these differences in Mortgage Yields were decreasing in the early 20th
century, they suprisingly remained stable after World War II. The paper
explores two main explanations for this phenomenon:\
\
**1.** Differences in investment value due to risk, terms, and
liquidity.\
**2.** Market imperfections such as legal barriers and information
gaps.\
\
The data used in this study comes from the Federal Home Loan Bank Board,
which contains interest rates and fees in 18 SMSAs (Standard
Metropolitan Statistical Areas). The findings suggest that distance from
major financial centers, risk levels, and local demand for savings
significantly affect Mortgage Yields. However, market structure and
overall savings levels play a lesser important role.\
\
The aim of this report is to analyze the data and develop a predictive
model to predict Mortgage Yield (`mortYld` in %) based on 6 explanatory
variables:\
\
- **X1:** Loan-to-Mortgage Ratio, in % → High values indicate low down
payments.\
- **X2:** Distance from Boston, in miles → Measures regional proximity
to financial centers.\
- **X3:** Savings per New Unit Built, in \$ → Indicator of regional
credit demand.\
- **X4:** Savings per Capita, in \$ → Measures local savings levels
(credit supply).\
- **X5:** Population Increase, 1950-1960, in % → Proxy for housing
demand growth.\
- **X6:** Percentage of First Mortgages from Inter-Regional Banks, in %
→ Indicator of external financing reliance.

# Exploratory Data Analysis (EDA)

## Load Data and Libraries

```{r global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r}
# Load necessary libraries
library(ggplot2)
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(xtable)
library(MASS)
library(car)
library(magrittr)
library(patchwork)
library(gridExtra)
library(grid)
library(reshape2)
library(broom)
library(xtable)

# Read data
df <- read.csv("mYield.csv")

# Rename columns for easier reference
colnames(df) <- c("smsa", "mortYld", "X1", "X2", "X3", "X4", "X5", "X6")

# Display first few rows
kable(head(df), caption = "First few rows of the dataset") %>%
  kable_styling(font_size = 8, latex_options = "hold_position")
```

```{r, include=FALSE}
# Count missing values
colSums(is.na(df))

```

Here is a display of the first few rows of the dataset. Each SMSA is described by
its Mortgage Yield (`mortYld`) as the dependent variable and six explanatory
variables (X1 to X6). All data consist of numerical values and quick checking confirms that there are no missing values in any region.

## Univariate Analysis

### Summary Statistics

```{r}
# Summary Statistics
summary_stats <- summary(df[, colnames(df) != "smsa"])
kable(summary_stats, caption = "Summary Statistics of Variables") %>%
  kable_styling(font_size = 8, latex_options = "hold_position")
```

Through this summary, we already observe that Mortgage Yields don’t vary
much across regions. Most values are between 5.2% and 6.2%, suggesting
relatively stable Mortgage rates.

Loan-to-Mortgage Ratios (X1) are concentrated in between 67% and 78.1%,
indicating relatively consistent lending practices across regions.
Distance from Boston (X2) has a vast range (0–3162 miles), highlighting
geographical diversity and potential financial access disparities.
Savings per New Unit Built (X3) and Savings per Capita (X4) are
characterized by means bigger than medians, representing right-skewed
distributions, thus suggesting regional imbalances in credit demand and
supply/in housing affordability across regions. Population Increase (X5)
from 1950 to 1960 varies widely (7.5–88.9%), reflecting differing
housing market pressures. Lastly, Percentage of First Mortgages from
Inter-Regional Banks (X6) spans from 2.0% to 51.3%, meaning that some
areas depend heavily on external financing while others rely more on
local institutions.

### Graphical Representation

```{r, fig.width=13, fig.height=5}

# Sort data by decreasing Mortgage Yield
df <- df %>% arrange(desc(mortYld))

# Define bar chart (corrected and stored as bar_plot)
bar_plot <- ggplot(df, aes(x = mortYld, y = reorder(smsa, mortYld))) + 
  geom_bar(stat = "identity", fill = "blue") +
  coord_cartesian(xlim = c(4.5, 6.5)) +
  labs(title = "Mortgage Yield by SMSA", x = "Mortgage Yield (%)", y = "SMSA (City)") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 9),
    plot.title = element_text(hjust = 0.5)
  )

# Define variable names and colors
var_names <- c("Loan-to-Mortgage Ratio", "Distance from Boston", 
               "Savings per New Unit Built", "Savings per Capita", 
               "Population Increase", "First Mortgage from Inter-Regional Banks")
hist_colors <- c("red", "blue", "green", "purple", "orange", "cyan")

# Create histogram plots
hist_plots <- lapply(1:6, function(i) {
  ggplot(df, aes_string(x = paste0("X", i))) +
    geom_histogram(fill = hist_colors[i], color = "black", bins = 8) +
    labs(title = var_names[i], x = NULL, y = "Frequency") +
    theme_minimal(base_size = 10)
})

# Combine histograms in 2x3 grid
hist_grid <- (hist_plots[[1]] | hist_plots[[2]] | hist_plots[[3]]) /
             (hist_plots[[4]] | hist_plots[[5]] | hist_plots[[6]])

# Combine bar chart and histograms side-by-side
final_plot <- bar_plot + hist_grid + plot_layout(widths = c(1.3, 2))

# Print final plot
final_plot
```

With deeper analysis, although the variation across SMSAs is small, we
see that regional differences still exist in Mortgage Yields, possibly
due to economic factors like savings, loan terms, and regional banking
practices.

The histograms confirm the distribution of the explanatory variables:

The Loan-to-Mortgage Ratio (X1) shows low variance with most values
concentrated between 67% and 80%, possibly indicating limited
variability across regions.
Distance from Boston (X2) displays a wide and almost homogeneous distribution, reflecting
substantial geographic spread among SMSAs.
Savings per New Unit Built (X3) and Savings per Capita (X4) both exhibit
right-skewed distributions, suggesting that a few cities have notably
higher savings levels.
Population Increase (X5) is also highly right-skewed with one major outlier (increase of ~25%),
indicating that most regions had moderate growth, while a few experienced
rapid expansion.
Finally, the percentage of First Mortgages from Inter-Regional Banks
(X6) is also right-skewed, with most cities relying minimally on
external financing and a few showing heavy dependence.

Overall, the data suggests regional variation in housing finance
conditions, credit accessibility, and Mortgage market dynamics.

## Bivariate Numerical Analysis

### Association Analysis

```{r, fig.width=4, fig.height=4}
# Réinitialisation des paramètres graphiques
par(mfrow = c(1, 1))
par(mai = c(0.3, 0.3, 0.5, 0.3))  # marges réduites : bas, gauche, haut, droite
par(oma = c(0, 0, 0, 0))          # aucune marge extérieure

# Sélection des variables numériques
selected_vars <- df[, c("mortYld", "X1", "X2", "X3", "X4", "X5", "X6")]

# Affichage du scatterplot matrix
pairs(selected_vars,
      main = "Association Matrix of Variables and mortYld",
      cex.main = 0.9,       # taille du titre réduite
      col = "blue",
      pch = 19)

```

The Association Matrix provides a quick visual assessment of bivariate relationships (how each variable relates to the others and `mortYld`), of types of associations among predictors (if a relationship looks linear, curved or weak, as well as positive or negative), and of outlier presence. It complements numerical analyses like the correlation matrix and VIF.

We can see that most of the plots are random dispersion, while some are
linear, and some are curved. X3 seems to be positively associated with
X4 and negatively with X5. X2 and X3 seem negatively exponentially associated.
X6 seems to be negatively associated with X3.

Let's take a closer look into the Association Matrix, regarding the
relationship between Mortgage Yield (%) and the explanatory variables
(x-axis), representing the first row in the precedent figure.

```{r, fig.width=10, fig.height=5}
# Set up a 2x3 grid for multiple scatter plots
par(mfrow = c(2, 3))

# Define variable names
var_names <- c("Loan-to-Mortgage Ratio", "Distance from Boston", 
               "Savings per New Unit Built", "Savings per Capita", 
               "Population Increase", "First Mortgage from Inter-Regional Banks")

# Define colors for better visualization
point_colors <- c("red", "blue", "green", "purple", "orange", "cyan")

# Loop through each variable (X1 to X6) and create a scatter plot with mortYld on the Y-axis
for (i in 1:6) {
  plot(df[[paste0("X", i)]], df$mortYld,
       main = paste(var_names[i], "vs Mortgage Yield"),
       xlab = var_names[i], ylab = "Mortgage Yield (%)",
       col = point_colors[i], pch = 19)
}

# Reset plotting parameters to default
par(mfrow = c(1, 1))

```
\
As the Loan-to-Mortgage Ratio (X1) increases, the Mortgage Yield increases. This suggests a positive correlation, and that higher
Loan-to-Mortgage Ratios (more borrowed money relative to the property
value) are associated with higher Mortgage Yields. Distance from Boston (X2) reveals a positive correlation with `mortYld`. Boston represents a major financial center with surplus capital. Regions
further from Boston might have higher Yields. Savings per New Unit Built (X3) seems to be negatively correlated with `mortYld`. This indicates that areas with more savings dedicated to new
construction have better access to local financing, resulting in lower Mortgage Yields. Savings per Capita (X4)'s influence is less distinguishable but appears to be a weak negative correlation or a random dispersion. Population Increase (X5) shows a positive association which can be seen as a square-root relationship. High population growth may imply
higher demand for housing, increasing Mortgage Yields due to heightened competition for available funds. We can observe a potential outlier at the right side of the plot. The Percentage of First Mortgages from Inter-Regional Banks (X6)'s variation shows no clear trend. It seems like the reliance on external financing does not significantly influence Mortgage Yields.\
\
**To resume:**\
- X1, X2 and X5 seem to be the most influential variables positively correlated with Mortgage Yield.\
- X3 is the most influential variable negatively correlated with Mortgage Yield.\
- X6 shows moderate positive influence on Mortgage Yield.\
- X4 variable shows a weak relationship with Mortgage Yields.\

These observations support the findings of Schaaf (1966) stating that distance
from financial centers, risk factors, and local demand for savings
contribute to Mortgage Yield variations.\

### Correlation Analysis

```{r, fig.width=4, fig.height=4}
library(reshape2)

# Compute correlation matrix and reshape
cor_matrix <- cor(df[, c("mortYld", "X1", "X2", "X3", "X4", "X5", "X6")])
melted_cor <- melt(cor_matrix)

# Plot
ggplot(data = melted_cor, aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 3) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name = "value") +
  coord_fixed() + 
  theme_minimal(base_size = 8) +
  theme(
    plot.title = element_text(size = 10, face = "bold", hjust = 0.5, margin = margin(b = 5)),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(5, 5, 5, 5)
  ) +
  labs(title = "Correlation Heatmap")

```

Now, let's take a look at the correlations between each variable and confirm our previous observations.
\
- X3 is are strongly positively correlated to X4 (0.77) and moderately negatively correlated to X2 (-0.64), X5 (-0.63) and X6 (-0.56). This confirms our observations with the Association Matrix.\
- X1 and X2 exhibit strong positive correlation with `mortYld`, while X5 shows moderate positive correlation, and X3 a
strong negative one. X6 shows moderate positive correlation with `mortYld` as well. X4 shows only weak correlation with `mortYld`. In consequence, we can confirm our previous statements about the scatter-plots.

We can then think about removing one of the highly correlated
predictors, to see if multicollinearity affects the regression model. However, these correlations only indicate if two variables are linearly associated. Thus, a low value doesn't necessarily mean that the variables are not correlated in another way.


------------------------------------------------------------------------

# Model Fitting

In this analysis, all predictors are continuous variables and each
observation corresponds to a unique SMSA. Since the dataset contains no
grouping or categorical factors with unequal group sizes, this is a
standard multiple regression model with one observation per row.
Therefore, the design is not factorial and does not involve unbalanced
group structures. As a result, the order in which predictors are entered
into the lm() function does not influence the coefficient estimates,
F-tests, or model interpretation.

## Pairwise Simple Regressions

```{r, fig.width=10, fig.height=2}
# Pairwise Simple Regressions
models_simple <- lapply(1:6, function(i) {
  formula <- as.formula(paste("mortYld ~ X", i, sep = ""))
  lm(formula, data = df)
})

simple_summaries <- lapply(models_simple, summary)

data.frame(
  Predictor = paste0("X", 1:6),
  R_squared = sapply(simple_summaries, function(m) round(m$r.squared, 3)),
  p_value = sapply(simple_summaries, function(m) round(coef(m)[2, 4], 4))
) %>%
  kable(caption = "Simple Linear Regressions: $R^2$ and p-values", escape = FALSE, booktabs = TRUE) %>%
  kable_styling(font_size = 8, latex_options = "hold_position")
```

The table summarizes the individual linear relationships between each predictor (X1–X6) and Mortgage Yield, assuming all other variables remain constant, based on simple linear regression.\
- X1 is the strongest and most significant predictor of Mortgage Yield, explaining 65.4% of its variance (p \< 0.0001).\
- X2 and X3 also show strong and significant linear associations ($R^2$ = 0.546 and 0.517, respectively with p \< 0.001 and p \< 0.001).\
- X5 and X6 show moderate yet significant associations ($R^2$ = 0.419 and 0.346, respectively with p \< 0.005 and p \< 0.05).\
- X4 doesn't exhibit a significant linear relationship with Mortgage Yield ($R^2$ = 0.049, p = 0.3763), suggesting it may not be a strong individual linear predictor.

This preliminary analysis indicates that variables X1, X2, and X3 may be the most promising candidates for predicting Mortgage Yield in a multivariate linear model.

## Null Model vs Full Model Comparison

```{r}
null_model <- lm(mortYld ~ 1, data = df)
full_model <- lm(mortYld ~ X1 + X2 + X3 + X4 + X5 + X6, data = df)

#anova(null_model, full_model)

# Résumé au format tableau propre
anova_table <- tidy(anova(null_model, full_model))

# Affichage du tableau formaté
kable(anova_table, caption = "Comparison of Null and Full Model (ANOVA)") %>%
  kable_styling(font_size = 8, latex_options = c("striped", "hold_position"))

```

The ANOVA comparison between the null model (intercept-only) and the full model (including all predictors), reveals that the full model better explains the Mortgage Yield, as shown by the significant F-statistic and p-value (p \< 0.001). This indicates that at least one of the predictors is significantly related to Mortgage
Yield, and is useful for improving the model.

```{r}
# Fit linear regression model (full)
model <- lm(mortYld ~ X1 + X2 + X3 + X4 + X5 + X6, data = df)

# Model summary
#summary(model)

# Résumé du modèle au format propre (dataframe)
model_summary <- tidy(model)

# Affichage du tableau dans le PDF
kable(model_summary, 
      caption = "Summary of Full Linear Model", 
      booktabs = TRUE, 
      digits = 4, 
      format = "latex", 
      escape = TRUE) %>%
  kable_styling(font_size = 8, latex_options = "hold_position") 

```
```{r}
### 2. Model Fit Statistics Table

# Extraire les statistiques de modélisation
model_stats <- glance(model)[, c("r.squared", "adj.r.squared", "sigma", "statistic","df", "p.value")]

# Renommer les colonnes en LaTeX-safe
colnames(model_stats) <- c(
  "R\\textsuperscript{2}", 
  "Adjusted R\\textsuperscript{2}", 
  "Residual Std. Error", 
  "F-statistic", 
  "DF",
  "p-value"
)

# Affichage du tableau
kable(model_stats,
      caption = "Fit Statistics of Full Linear Model", 
      booktabs = TRUE, digits = 4, format = "latex", escape = FALSE) %>%
  kable_styling(font_size = 8, latex_options = "hold_position", position = "center")


```

The Full model explains ~87% of the variance in Mortgage Yield, and 80% after adjusting for the number of predictors, which highlights a strong fit. The Residual Standard Error (?) is low, and the overall model is statistically significant, with a very low p-value (p \< 0.001). Once again, it means that at least one term contributes significantly to explaining the variation in `mortYld`.

The intercept appears to be strongly significant to fit the model (p \< 0.001). On the other hand, most of the variables do not show statistically significant individual contributions: only X1 and X3 show weak significance (p $\approx$ 0.05), while the other variables, X2, X5 and X6, do not show significant individual effects. This suggests that a reduced model may be more appropriate.

We end up with : `mortYld` = 4.2852 + 0.0203\*X1 + 0.0\*X2 - 0.0016\*X3 + 0.0002\*X4 + 0.0013\*X5 + 0.0002\*X6

## Make stepwise regression to select the best model

```{r}
# 1. Stepwise AIC Steps Table

# Stepwise AIC with silent output
step_trace <- capture.output(
  step_model <- stepAIC(model, direction = "both", trace = TRUE)
)

# Build Stepwise AIC Trace Manually
step_trace_df <- data.frame(
  Step = c("Start", "Step 1", "Step 2", "Step 3"),
  Model = c(
    "X1 + X2 + X3 + X4 + X5 + X6",
    "X1 + X2 + X3 + X4 + X5",
    "X1 + X3 + X4 + X5",
    "X1 + X3 + X4"
  ),
  RSS = c(0.1098, 0.1099, 0.1109, 0.1159),
  AIC = c(-77.79, -79.77, -81.61, -82.81)
)

# Display with kable
kable(step_trace_df, 
      caption = "Stepwise AIC Steps", 
      booktabs = TRUE, 
      format = "latex", 
      digits = 4) %>%
  kable_styling(font_size = 8, latex_options = "hold_position", position = "center")

```

```{r}
#  2. Model Coefficients Table
step_summary <- tidy(step_model)

kable(step_summary, 
      caption = "Coefficients of Final Stepwise Model", 
      booktabs = TRUE, digits = 4, format = "latex", escape = TRUE) %>%
  kable_styling(font_size = 8, latex_options = "hold_position", position = "center")

```

```{r}
# 3. Fit Statistics of Stepwise Model

# Extraire les stats globales
step_glance <- glance(step_model)[, c("r.squared", "adj.r.squared", "sigma", "statistic", "df", "p.value")]
colnames(step_glance) <- c("R²", "Adjusted R²", "Residual Std. Error", "F-statistic", "DF", "p-value")

# Convertir en data.frame pour mise en forme fine
step_glance <- as.data.frame(step_glance)

# Appliquer un arrondi classique sauf pour la p-value
step_glance$`R²` <- round(step_glance$`R²`, 4)
step_glance$`Adjusted R²` <- round(step_glance$`Adjusted R²`, 4)
step_glance$`Residual Std. Error` <- round(step_glance$`Residual Std. Error`, 4)
step_glance$`F-statistic` <- round(step_glance$`F-statistic`, 4)
step_glance$`DF` <- as.integer(step_glance$`DF`)  # entier
step_glance$`p-value` <- formatC(step_glance$`p-value`, format = "e", digits = 2)

# Afficher en tableau LaTeX
kable(step_glance,
      caption = "Fit Statistics of Stepwise Model",
      booktabs = TRUE, format = "latex", escape = TRUE) %>%
  kable_styling(font_size = 8, latex_options = "hold_position", position = "center")


```

The Stepwise regression process identifies X1, X3, and X4 as the most
significant predictors of Mortgage Yield, constituting the final model.

It is interesting to note that X4 appears among the 3 most significant predictors although it shows very weak correlation in the Correlation Matrix. Multiple regression measures the effect of each variable while holding all others constant. As X4 has very strong correlation with X3 (0.77), holding X3 can make the unique contribution of X4 clearer.

The final Stepwise model explains approximately 83.4% of the variance in Mortgage Yield using only these three predictors. The AIC doesn't increases a lot when keeping more predictors, meaning that even if these predictors can still be statistically valid to keep, they are not so useful to the model. Though the final model is simpler, it explains the data just as well or better than more complex models. The Residual Standard Error (0.091) is low, and the overall model is highly significant (p \< 0.001), indicating a good fit.

We end up with : `mortYld` = 4.223 + 0.02229\*X1 - 0.001863\*X3 + 0.0002249\*X4

Let's now try a model with 2-way interactions.

```{r}
# Fit model
interaction_model <- lm(mortYld ~ X1 + X3 + X4 + X1:X3 + X1:X4 + X3:X4, data = df)

### 1. Coefficients Table
interaction_coef <- tidy(interaction_model)

kable(interaction_coef, 
      caption = "Coefficients of Interaction Model", 
      booktabs = TRUE, digits = 4, format = "latex", escape = TRUE) %>%
  kable_styling(font_size = 8, latex_options = "hold_position", position = "center")

```


```{r}
### 2. Model Fit Statistics Table

# Extraire les statistiques de modélisation
interaction_stats <- glance(interaction_model)[, c("r.squared", "adj.r.squared", "sigma", "statistic", "df", "p.value")]

# Renommer les colonnes en LaTeX-safe
colnames(interaction_stats) <- c(
  "R\\textsuperscript{2}", 
  "Adjusted R\\textsuperscript{2}", 
  "Residual Std. Error", 
  "F-statistic", 
  "DF",
  "p-value"
)

# Affichage du tableau
kable(interaction_stats,
      caption = "Fit Statistics of Interaction Model", 
      booktabs = TRUE, digits = 4, format = "latex", escape = FALSE) %>%
  kable_styling(font_size = 8, latex_options = "hold_position", position = "center")


```

The 2-way Interactions model, which is more complex than the Stepwise model, explains approximately 79.9% of the variance in Mortgage Yield. The Residual Standard Error (0.1002) is low, and the overall model is highly significant (p \< 0.001), indicating that at least one of the terms has a significant influence on Mortgage Yield.

None of the variables show statistically significant individual contributions: only the intercept appears to be moderately significant to fit the model (p \< 0.05). This suggests that a reduced model may be more appropriate.

We end up with : `mortYld` = 5.3710 + 0.0069\*X1 - 0.0001\*X3 - 0.0009\*X4 + 0.0\*X1:X3 + 0.0\*X1:X4 + 0.0\*X3:X4

We decided not to include a 3-way Interactions model in our analysis.
Given the small sample size (18 observations), adding high-order
interactions would significantly reduce degrees of freedom and increase
the risk of overfitting. Moreover, 3-way interactions are often
difficult to interpret meaningfully.

## Model Comparison

```{r, fig.width=10, fig.height=2}
# Create comparison table
model_metrics <- data.frame(
  Model = c("Full Model", "Stepwise Model", "2-Way Interaction Model"),
  R2 = c(
    summary(full_model)$r.squared,
    summary(step_model)$r.squared,
    summary(interaction_model)$r.squared
  ),
  Adj_R2 = c(
    summary(full_model)$adj.r.squared,
    summary(step_model)$adj.r.squared,
    summary(interaction_model)$adj.r.squared
  ),
  AIC = c(
    AIC(full_model),
    AIC(step_model),
    AIC(interaction_model)
  ),
  Residual_SE = c(
    summary(full_model)$sigma,
    summary(step_model)$sigma,
    summary(interaction_model)$sigma
  ),
  F_statistic = c(
    summary(full_model)$fstatistic[1],
    summary(step_model)$fstatistic[1],
    summary(interaction_model)$fstatistic[1]
  )
)

# Display table
kable(model_metrics, digits = 3, caption = "Comparison of Model Performance Metrics") %>%
  kable_styling(font_size = 8, full_width = FALSE, position = "center")

```

The **Stepwise model** offers the best trade-off between simplicity and performance: it has the lowest AIC (~29.7), demonstrating the best model fit among the three. Despite having a slightly lower R² than the Full and 2-ways Interactions model, it achieves the highest Adjusted R². It also has the lowest Residual Standard Error (0.091) and the highest F-statistic (~29.5). This confirms the overall model significance and parsimony.

```{r}
# Compare with ANOVA and convert to data frame
anova_result <- anova(step_model, interaction_model)

# Ajouter les noms des modèles pour plus de clarté
anova_df <- as.data.frame(anova_result)
anova_df$Model <- c("Stepwise model", "Interaction model")

# Réorganiser les colonnes
anova_df <- anova_df[, c("Model", "Res.Df", "RSS", "Df", "Sum of Sq", "F", "Pr(>F)")]

# Affichage avec kable
kable(anova_df,
      caption = "ANOVA Comparison: Stepwise vs Interactions Model",
      booktabs = TRUE, digits = 4, format = "latex", escape = TRUE) %>%
  kable_styling(font_size = 8, latex_options = "hold_position", position = "center")

```

An ANOVA is then conducted to confirm that including 2-way Interactions
terms do not significantly improve the model fit. The test yields an
F-statistic of 0.18 and a p-value of 0.91, indicating that the
additional interaction terms do not meaningfully reduce the residual
variance.

As a result, the simpler model with only main effects (X1,
X3, and X4) truly provides the best fit, as it also offers comparable explanatory power and better
interpretability.

------------------------------------------------------------------------

# Model assumptions and Diagnostics

## Independence evaluation

```{r, fig.width=7.5, fig.height=2.5}
par(mfrow = c(1, 3),        # 1 row, 3 columns
    mar = c(3.5, 3.5, 2, 1), # Margins: bottom, left, top, right
    mgp = c(2, 0.6, 0),     # Axis label position
    cex.main = 0.8,         # Title size
    cex.lab = 0.8,          # Axis label size
    cex.axis = 0.7)         # Tick mark size

# Residuals vs. Fitted Values Plot
plot(fitted(model), resid(model),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals",
     pch = 19, col = "blue")
res_fit_lm <- lm(resid(model) ~ fitted(model))  # Linear model
abline(res_fit_lm, col = "red", lwd = 2)  # Linear trend line
coef_res_fit <- coef(res_fit_lm)
text(min(fitted(model)), max(resid(model)), 
     labels = paste0("y = ", round(coef_res_fit[2], 3), "x + ", round(coef_res_fit[1], 3)), 
     pos = 4, col = "red", cex = 0.8)

# Residuals vs. Observation Order Plot
plot(resid(model), type = "p",
     main = "Residuals vs Observation Order",
     xlab = "Observation Index", ylab = "Residuals",
     pch = 19, col = "darkgreen")
res_obs_lm <- lm(resid(model) ~ c(1:length(resid(model))))  # Linear model
abline(res_obs_lm, col = "red", lwd = 2)  # Linear trend line
coef_res_obs <- coef(res_obs_lm)
text(1, max(resid(model)), 
     labels = paste0("y = ", round(coef_res_obs[2], 3), "x + ", round(coef_res_obs[1], 3)), 
     pos = 4, col = "red", cex = 0.8)

# Scale-Location Plot (Spread-Location)
plot(fitted(model), sqrt(abs(resid(model))),
     main = "Scale-Location Plot",
     xlab = "Fitted Values", ylab = "√|Residuals|",
     pch = 19, col = "purple")
scale_loc_lm <- lm(sqrt(abs(resid(model))) ~ fitted(model))  # Linear model
abline(scale_loc_lm, col = "red", lwd = 2)  # Linear trend line
coef_scale_loc <- coef(scale_loc_lm)
text(min(fitted(model)), max(sqrt(abs(resid(model)))),
     labels = paste0("y = ", round(coef_scale_loc[2], 3), "x + ", round(coef_scale_loc[1], 3)), 
     pos = 4, col = "red", cex = 0.8)

# Reset plot settings
par(mfrow = c(1, 1))

```

The `Residuals VS Fitted Values` plot shows that the residuals are randomly scattered around 0, with no clear pattern. This suggests that the assumptions of linearity and constant error variance are reasonably met. Secondly, the slightly negative but close to zero slope (-0.286) in the `Scale-Location Plot` indicates that the spread of residuals is almost constant across fitted values. This is a sign that our model doesn't suffer from heteroscedasticity and is likely a good fit: homoscedasticity seems therefore satisfied.

The `Residuals VS Observation Order` plot helps us conclude that there is no consistent trend: the independence of residuals is verified, as they are not correlated with the order of observations.

## Multicolinearity diagnostic

```{r, fig.width=8, fig.height=3}
# Compute Variance Inflation Factor (VIF)
vif_values <- vif(step_model)
kable(as.data.frame(vif_values), caption = "Variance Inflation Factors (VIF)") %>% 
  kable_styling(font_size = 8, latex_options = "hold_position")

```

As all variables have a VIF value under 5, it means that they don't cause problematic multicollinearity in the final model
and that none of them should be eliminated. This confirms our choice of keeping X3 and X4: even if they showed a high
correlation coefficient (0.77), these variables still provide enough unique, non-redundant information to justify keeping them in the model.

## Normality Check

```{r, fig.width=3, fig.height=3}
# Q-Q Plot with equal axis limits
par(pty = "s", 
    mar = c(4, 3.5, 2, 1),    # ← marge gauche réduite à 3.5
    mgp = c(2, 0.5, 0),       # position des labels des axes
    cex.main = 0.8,        # title size
    cex.lab = 0.8,         # axis label size
    cex.axis = 0.7)        # axis ticks

qqnorm(resid(step_model), 
       main = "Q-Q plot of Residuals", 
       xlab = "Theoretical Quantiles", 
       ylab = "Sample Quantiles",
       pch = 19, col = "steelblue",
       xlim = c(-2, 2), ylim = c(-0.2, 0.2))
qqline(resid(step_model), col = "red")
```

The `Q-Q plot of Residuals` indicates that most of the points align closely with the 45-degree line, suggesting that the residuals are approximately normally distributed. However, six points deviate at the extremes of the theoretical quantiles, which indicates the presence of potential outliers or heavy tails in the distribution. In a dataset with just 18 observations, small deviations in the Q-Q plot are usual. Outliers or deviations are common in such a small sample size and do not automatically suggest a violation of normality. 

# Conclusions
\
- The final estimated model is :
`mortYld` = 4.223 + 0.02229\*X1 - 0.001863\*X3 + 0.0002249\*X4\
where X1 is the Loan-to-Mortgage Ratio, X3 is the Savings per New Unit Built, and X4 is the Savings per Capita.\
- The analysis thus shows that Loan-to-Mortgage Ratio, Savings per New Unit Built, and Savings per Capita collectively have a significant impact on Mortgage Yield. Mortgage Yield is positively influenced by the Loan-to-Mortgage Ratio (X1), suggesting that when the loan is higher compared to the mortgage, lenders may see better returns on their investment. On the other hand, Mortgage Yield is negatively impacted by Savings per New Unit Built (X3). When there is more capital saved for new construction, there may be less reliance on mortgages, potentially leading to lower returns on those mortgages. Lastly, Savings per Capita (X4) has a positive, though small, effect on the Mortgage Yield. As individuals save more money, it may lead to a slight increase in the return on mortgages, possibly because higher savings per capita can signal a more financially stable environment for lenders, leading to better mortgage performance.\
- While the assumptions of linear regression are generally satisfied, there are some minor deviations.\
- The model shows strong predictive performance, accounting for 83.4% of the variance in Mortgage Yield, with homoscedasticity nearly achieved.\
- Future improvements could include exploring additional predictors, testing for non-linear relationships, or refining the model to better capture any residual heteroscedasticity.

