---
title: "Predicting Mortgage Yield using Regression Analysis"
author: "Group 42"
date: "`r Sys.Date()`"
output: pdf_document
geometry: margin=2.5cm
fontsize: 12pt
header-includes:
  - \usepackage{graphicx}
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{Clara Delandre, Majandra Garcia, Paola Biocchi, Coline Leteurte}
  - \fancyfoot[C]{\thepage}
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

Mortgage Yield (`mortYld`) is predicted based on economic and
demographic factors. This study examines how **various economic and
geographic factors influence mortgage yield (%)**, using data from
historical mortgage reports. The aim is to develop a predictive model
for mortgage yield based on six explanatory variables:

-   **X1:** Loan-to-Mortgage Ratio (high values indicate low down
    payments)
-   **X2:** Distance from Boston (miles)
-   **X3:** Savings per new unit built
-   **X4:** Savings per capita
-   **X5:** Population increase (1950-1960, %)
-   **X6:** First mortgage from inter-regional banks (%)

------------------------------------------------------------------------

# Exploratory Data Analysis (EDA)

## Load Data and Libraries

```{r}
# Load necessary libraries
library(ggplot2)
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(xtable)
library(MASS)
library(car)
library(magrittr)

# Read data
df <- read.csv("myield.csv")

# Rename columns for easier reference
colnames(df) <- c("smsa", "mortYld", "X1", "X2", "X3", "X4", "X5", "X6")

# Display first few rows
kable(head(df), caption = "First few rows of the dataset") %>% kable_styling()
```

```{r}
# Count missing values
colSums(is.na(df))

```

## Summary Statistics

```{r}
# Summary Statistics
summary_stats <- summary(df)
kable(summary_stats, caption = "Summary Statistics of Variables") %>% kable_styling()
```

## Correlation Analysis

```{r}
# Correlation Matrix
# Compute correlation matrix
cor_matrix <- round(cor(df[,3:8]), 2)  # Excluding 'smsa' and 'mortYld'
kable(cor_matrix, caption = "Correlation Matrix") %>% kable_styling()
```

## Visualizations

```{r}
# Histogram of Mortgage Yield
ggplot(df, aes(x = mortYld)) + 
  geom_histogram(binwidth = 0.2, fill = "blue", alpha = 0.6) +
  labs(title = "Distribution of Mortgage Yield", x = "Mortgage Yield (%)", y = "Count")

# Scatter Plot: Mortgage Yield vs Loan/Mortgage Ratio
ggplot(df, aes(x = X1, y = mortYld)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Mortgage Yield vs Loan/Mortgage Ratio", 
       x = "Loan/Mortgage Ratio", y = "Mortgage Yield (%)")
```

# Model Fitting

```{r}
# Fit linear regression model
model <- lm(mortYld ~ X1 + X2 + X3 + X4 + X5 + X6, data = df)

# Model summary
summary(model)

```

## Make stepwise regression to select the best model

By removing some of the variables

```{r}
# Stepwise regression (both directions)
step_model <- stepAIC(model, direction = "both")
summary(step_model)
```

# Model assumptions and Diagnostics

## Multicolinearity Check using VIF

a VIF \> 5 indicates possible multicollinearity that can be problematic
: we are not able to identify the variables associated with this
collinearity but we will have to do it next.

```{r}
# Compute Variance Inflation Factor (VIF)
vif_values <- vif(model)
kable(as.data.frame(vif_values), caption = "Variance Inflation Factors (VIF)") %>% kable_styling()

```

## Residuals Analysis

Random scatter indicates good assumption of homeoscedasticity. If we can
distinguish a clear pattern, then we have potential heteroscedasticity
issue.

```{r}
# Residuals vs Fitted Plot
ggplot(data.frame(fitted = fitted(model), residuals = resid(model)), aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Plot", x = "Fitted Values", y = "Residuals")
```

## Normality Check

If points lie on 45 degrees line, it means the residuals are normally
distributed. If we can see a curved pattern, then the normality
assumption is violated

```{r}
# QQ-Plot of Residuals
qqnorm(resid(model))
qqline(resid(model), col = "red")
```

# Final estimated Model

```{r}
coeffs <- round(coef(model), 2)
kable(as.data.frame(coeffs), caption = "Final Model Coefficients") %>% kable_styling()
```

# Conclusions

```         
•   The analysis showed that [mention significant predictors] have a strong relationship with mortgage yield.
•   The assumptions of linear regression were [state if met or violated].
•   The model provides [good/poor] predictive accuracy based on [R² and residual analysis].
•   Future improvements could involve [mention possible improvements like transformations, additional predictors, etc.].
```
