---
title: "Predicting Mortgage Yield using Regression Analysis"
author: "Group 42"
date: "`r Sys.Date()`"
output: pdf_document
geometry: margin=2.5cm
fontsize: 12pt
header-includes:
  - \usepackage{graphicx}
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \usepackage{fancyhdr}
  - \usepackage{ragged2e}
  - \justifying
  - \pagestyle{fancy}
  - \fancyhead[L]{Clara Delandre, Majandra Garcia, Paola Biocchi, Coline Leteurte}
  - \fancyfoot[C]{\thepage}
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

The study of A. H. Schaaf, 1966, "Regional Differences in Mortgage
Financing Costs" investigates the existence and causes of regional
differences in mortgage financing costs in the United States. These
differences in mortgage yields have decreased in the early 20th century,
however they remained stable after World War 2. The paper explores two
main explanations for this phenomenon:\
\
**1.** differences in investment value due to risk, terms, and
liquidity\
**2.** market imperfections such as legal barriers and information
gaps.\
\
The data used in this study comes from the Federal Home Loan Bank Board,
which contains interest rates and fees in 18 SMSAs (Standard
Metropolitan Statistical Areas). The findings suggest that distance from
major financial centers, risk levels, and local demand for savings
significantly affect mortgage yields. However, market structure and
overall savings levels play a lesser role.\
\
The aim of this report is to analyze the data and develop a predictive
model to predict Mortgage Yield (`mortYld`) based on 8 explanatory
variables:\
\
- **smsa:** Standard Metropolitan Statistical Areas (18) → Name of the
city/region.\
- **mortYld:** Mortgage Yield, in % → The percentage return on a
mortgage.\
- **X1:** Loan-to-Mortgage Ratio, in % → High values indicate low down
payments.\
- **X2:** Distance from Boston, in miles → Measures regional proximity
to financial centers.\
- **X3:** Savings per new unit built, in \$ → Indicator of regional
credit demand.\
- **X4:** Savings per capita, in \$ → Measures local savings levels
(credit supply).\
- **X5:** Population increase, 1950-1960, in % → Proxy for housing
demand growth.\
- **X6:** Percentage of first mortgages from inter-regional banks, in %
→ Indicator of external financing reliance.\

------------------------------------------------------------------------

# Exploratory Data Analysis (EDA)

## Load Data and Libraries

```{r}
# Load necessary libraries
library(ggplot2)
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(xtable)
library(MASS)
library(car)
library(magrittr)

# Read data
df <- read.csv("myield.csv")

# Rename columns for easier reference
colnames(df) <- c("smsa", "mortYld", "X1", "X2", "X3", "X4", "X5", "X6")

# Display first few rows
kable(head(df), caption = "First few rows of the dataset") %>% kable_styling()
```

```{r}
# Count missing values
colSums(is.na(df))

```

Here is a display of the data, on the first few rows of the dataset. It
contains mortgage yield (mortYld) as the dependent variable and six
variables (X1 to X6). smsa represents the Standard Metropolitan
Statistical Area which is the name of the city/region. We can observe
that all data are numerical values and there is no missing value for
each region.

## Univariate Analysis

### Summary Statistics

```{r}
# Summary Statistics
summary_stats <- summary(df)
kable(summary_stats, caption = "Summary Statistics of Variables") %>% kable_styling()
```

We can observe that each variable 18 observations corresponding to 18
different SMSAs. Through this summary, we can already observe mortgage
yields don’t vary much across regions. Most values are between 5.5% and
6.2%, suggesting relatively stable mortgage rates.

Loan-to-mortgage ratios (X1) are concentrated in between 70% and 78%,
with low variance. Savings per New Unit Built (X3) are characterized by
a mean bigger than the median, representing a right-skewed distribution
and then large disparities in housing affordability across regions.

-\> to complete

### Graphical Representation

```{r}
# Histogram of Mortgage Yield
ggplot(df, aes(x = mortYld)) + 
  geom_histogram(binwidth = 0.2, fill = "blue", alpha = 0.6) +
  labs(title = "Distribution of Mortgage Yield", x = "Mortgage Yield (%)", y = "Count")

# Create a named vector where mortgage yield values are associated with their respective SMSAs
df <- df %>% arrange(desc(mortYld))  

# Create a better bar chart
ggplot(df, aes(x = reorder(smsa, mortYld), y = mortYld)) + 
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() +  # Flip coordinates to make labels readable
  labs(title = "Mortgage Yield by SMSA", x = "SMSA (City)", y = "Mortgage Yield (%)") +
  theme_minimal()


# Set up a 2x3 plotting area
par(mfrow = c(2, 3))  

# Define variable names
var_names <- c("Loan-to-Mortgage Ratio", "Distance from Boston", 
               "Savings per New Unit Built", "Savings per Capita", 
               "Population Increase", "First Mortgage from Inter-Regional Banks")

# Define colors for better visualization
box_colors <- c("red", "blue", "green", "purple", "orange", "cyan")

# Loop through each variable (X1 to X6) and create a boxplot
for (i in 1:6) {
  boxplot(df[[paste0("X", i)]], 
          main = var_names[i], 
          col = box_colors[i], 
          ylab = "Value")
}

# Reset plotting parameters to default
par(mfrow = c(1, 1))
```

1st graph : doesn't seem to follows a normal distribution.

2nd graph : There is not a huge variation in mortgage yield across
SMSAs, as most bars are at similar heightsbut if we focus on 4 and 6%,
we see regional differences exist in mortgage yields, possibly due to
economic factors like savings, loan terms, and regional banking
practices.

3rd graph : There is a little variability for Loan-to-Mortgage Ratio,
which might suggest that this variable has limited predictive power in
explaining mortgage yield variation. X3 and X4 have right-skewed
distribution (longer upper whisker) meaning a few cities have very high
savings. It could indicate higher affordability and lower mortgage yield
in some regions. we can see there is at least an outlier for Population
Increase (X5)

## Bivariate Numerical Analysis

### Association Analysis

```{r}
# Set up a 2x3 grid for multiple scatter plots
par(mfrow = c(2, 3))

# Define variable names
var_names <- c("Loan-to-Mortgage Ratio", "Distance from Boston", 
               "Savings per New Unit Built", "Savings per Capita", 
               "Population Increase", "First Mortgage from Inter-Regional Banks")

# Define colors for better visualization
point_colors <- c("red", "blue", "green", "purple", "orange", "cyan")

# Loop through each variable (X1 to X6) and create a scatterplot against mortYld
for (i in 1:6) {
  plot(df[[paste0("X", i)]], df$mortYld,
       main = paste(var_names[i], "vs Mortgage Yield"),
       xlab = var_names[i], ylab = "Mortgage Yield (%)",
       col = point_colors[i], pch = 19)
}

# Reset plotting parameters to default
par(mfrow = c(1, 1))

```
Tell about positive, negative associations, weak, strong ? (week 2 slide 5)

### Correlation analysis

```{r}
# Correlation Matrix
# Compute correlation matrix
cor_matrix <- round(cor(df[,3:8]), 2)  # Excluding 'smsa' and 'mortYld'
kable(cor_matrix, caption = "Correlation Matrix") %>% kable_styling()
```
Numerical interpretation : LINEAR RELASHIONSHIPS

X3 (Savings per New Unit Built) and X4 (Savings per Capita) seem to be
strongly correlated (0.77). X2 (Distance from Boston) and X3 (Savings
per New Unit Built) have a high negative correlation (-0.64). X3
(Savings per New Unit Built) is also negatively correlated with X5
(Population Growth) at -0.63.

We can then think about removing one of the highly correlated
predictors, if multicollinearity affects the regression model.

numerical value of r : curve ? outliers? parallel lines? 

### Graphical : Heatmap and Scatter Plot

```{r}
library(ggplot2)
library(reshape2)

# Convert correlation matrix to long format for heatmap
melted_cor <- melt(cor_matrix)

# Reverse the order of the y-axis (Var2) so that X1 appears at the top-left
ggplot(data = melted_cor, aes(x=Var1, y=factor(Var2, levels=rev(unique(Var2))), fill=value)) + 
  geom_tile() +
  scale_fill_gradient2(low="blue", high="red", mid="white", 
                       midpoint=0, limit=c(-1,1), space="Lab") +
  scale_x_discrete(position = "top") +  # Move X-axis labels to top
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0)) + # Rotate labels for readability
  ggtitle("Correlation Heatmap")
```
### faire une dispersion matrix ? slide 28 cours 2

------------------------------------------------------------------------

# Model Fitting

```{r}
# Fit linear regression model
model <- lm(mortYld ~ X1 + X2 + X3 + X4 + X5 + X6, data = df)

# Model summary
summary(model)

```

## Make stepwise regression to select the best model

By removing some of the variables

```{r}
# Stepwise regression (both directions)
step_model <- stepAIC(model, direction = "both")
summary(step_model)
```

The stepwise regression process identified X1, X3, and X4 as the most
significant predictors of mortality yield, leading to the final model:\
mortYld = 4.223 + 0.02229.X1 − 0.001863.X3 + 0.0002249.X4\
Now we will test a model with 3-way interactions.

```{r}
#model with interactions
interaction_model <- lm(mortYld ~ X1 * X3 * X4, data = df)

# Summary of the model
summary(interaction_model)

```

The model complexity increased, but there was no significant improvement
in performance.\
Let's try a model with only 2-way interactions, and use a AIC step-wise
selection to select only relevant interactions

```{r}
reduced_interaction_model <- lm(mortYld ~ X1 + X3 + X4 + X1:X3 + X1:X4 + X3:X4, data = df)
summary(reduced_interaction_model)
step_interaction_model <- stepAIC(lm(mortYld ~ X1 * X3 * X4, data = df), direction = "both")
summary(step_interaction_model)

```

This model is worse then with all 3-way interactions.

------------------------------------------------------------------------

# Model assumptions and Diagnostics

## Multicolinearity Check using VIF

a VIF \> 5 indicates possible multicollinearity that can be problematic

:   we are not able to identify the variables associated with this
    collinearity but we will have to do it next.

```{r}
# Compute Variance Inflation Factor (VIF)
vif_values <- vif(model)
kable(as.data.frame(vif_values), caption = "Variance Inflation Factors (VIF)") %>% kable_styling()

```

## Residuals Analysis

Random scatter indicates good assumption of homeoscedasticity. If we can
distinguish a clear pattern, then we have potential heteroscedasticity
issue.

```{r}
# Residuals vs Fitted Plot
ggplot(data.frame(fitted = fitted(model), residuals = resid(model)), aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Plot", x = "Fitted Values", y = "Residuals")
```

## Normality Check

If points lie on 45 degrees line, it means the residuals are normally
distributed. If we can see a curved pattern, then the normality
assumption is violated

```{r}
# QQ-Plot of Residuals
qqnorm(resid(model))
qqline(resid(model), col = "red")
```

------------------------------------------------------------------------

# Final estimated Model

```{r}
coeffs <- round(coef(model), 2)
kable(as.data.frame(coeffs), caption = "Final Model Coefficients") %>% kable_styling()
```

------------------------------------------------------------------------

# Conclusions

```         
•   The analysis showed that [mention significant predictors] have a strong relationship with mortgage yield.
•   The assumptions of linear regression were [state if met or violated].
•   The model provides [good/poor] predictive accuracy based on [R² and residual analysis].
•   Future improvements could involve [mention possible improvements like transformations, additional predictors, etc.].
```
